---
title: "Prepare Datasets to MetaboAnalyst"
author: "David Garrido Rodríguez"
date: "`r Sys.Date()`"
output: html_document
---

# Script description:

The **aim** of this script is to prepare pre-annotated untargeted metabolomics data for use in MetaboAnalyst. It performs several tasks to achieve this:

1. Reads the raw data files from the input folder.
2. For each file, it performs data preprocessing steps such as removing duplicates metabolites, eliminating unidentified ones, and selecting desired columns (individual areas).
3. Merges the processed data frames based on file names to create dynamic tables.
4. Saves the merged data frames as CSV files for further analysis in MetaboAnalyst.

To run this script, you need to provide the input and output folder addresses, as well as an Excel file (.xlsx) containing the desired column (current names), new columns names, and groups for each sample. The structure of this file should follow the instructions provided in point 5 of the usage instructions.

**By performing these steps, the script ensures that the metabolomics data is properly cleaned and formatted, making it compatible with MetaboAnalyst, which is a powerful tool for metabolomics data analysis.**

--------------------------------------------------------------------------------

# USAGE INSTRUCTIONS:

1. The environment must be clear. If not, run this chunck:
```{r}
rm(list = ls())
```

2. Input data must be in a individual folder with the next name structure:
  **"Compounds_essayname_individual_characteristics.xlsx"**
    --> I.e. = Compounds_Global_noOTA_d2_15_S17:
      - Global = all noOTA dataframes
      - noOTA_d2_15_S17 = no production of OTA, day 2 of sampling, _Penicillium nordicum_ 15, culture medium (salchichón 17% glyc./90 a.w.)
      
  Please note that the part following "Compounds_" will determine the new data name and with which other dataset it is going to be merge. In this case, for example, the formatted data from Compounds_Global_noOTA_... and Compounds_Global_noOTA_... will be named "Global_merged."

3.The input data should be pre-annotated with the names in the first column (named as "Name") and should have the individual areas of each condition in columns named as 'Area: '.

4. Correspondences excel book must have the following columns but with the information that you need (Aditional_info is opcional). The columns 'name', 'formula', 'mz' and 'RT' must always have that new names:

  | Actual_name  |  New_name  | Group | 
  | ------------ | ---------- | ----- | 
  | Area: A1.raw | 2_1_15_S17 | noOTA |
  | Area: A2.raw | 2_2_15_S17 | noOTA |
  | Area: A3.raw | 2_3_15_S17 |  OTA  |
  | Area: A4.raw | 2_4_15_S17 |  OTA  |
  |    'm/z'     |     mz     |       |
  |   RT [min]   |     RT     |       |

5. You will need the following packages
```{r message=FALSE, warning=FALSE}
if (!require("tidyverse")) install.packages("tidyverse",repos ="https://cran.rstudio.com/")
library(tidyverse)
if (!require("tools")) install.packages("tools")
library(tools)
```

--------------------------------------------------------------------------------
# Set the input, output **folders** and the correspondences excel path here:
```{r}
# If you want to generate STATISTICAL ANALYSIS [ONE FACTOR] csv please write here "Yes" or "No":
input_folder <- "../../1_Data/Comparative_G/Untargeted_metabolomic_data/Individual_conditions"
output_folder <- "../../1_Data/Comparative_G/Untargeted_metabolomic_data/Ready_to_use"
Correspondences <- "../../1_Data/Comparative_G/Correspondences.xlsx"
```
Then, run the RScript by pressing Run/Restard R and Run All Chunks.
--------------------------------------------------------------------------------


--------------------------------------------------------------------------------
--------------------------------------------------------------------------------
--------------------------------------------------------------------------------

# Prepare dataset to MetaboAnalyst - STATISTICAL ANALYSIS [ONE FACTOR]:

## Import raw data and edit it

```{r}
files <- list.files(input_folder)
env_files <- character()

# Load correspondences dataset and verify that the necessary information is contained
correspondences <- readxl::read_xlsx(Correspondences)
  
if (!all(c("Current_name", "New_name") %in% colnames(correspondences))) {
  stop("'Current_name' and 'New_name' columns do not exit or have other names in the correspondences dataset you have provide. Check the dataset and rerun.")
}
if (!"Group" %in% colnames(correspondences)) {
  stop("'Group' column do not exit or have other name in the correspondences dataset you have provide. Check the dataset and rerun.")
}

# Clean data set and prepare them
for (i in files){
  
  # Read data files
  data_path <- file.path(input_folder, i)
  raw_data <- readxl::read_xlsx(data_path)
  ## Ensure data have the needed structure:
  if (!any(grepl("^Name$", names(raw_data), ignore.case = TRUE))) {
    stop(sprintf("The 'Name' column is not present in the dataset of the file '%s'.", i))
  } else if (!any(grepl("^Formula", colnames(raw_data), ignore.case = TRUE))) {
    stop(sprintf("The formula column is not present in the dataset of the file '%s'.", i))
  } else if (!any(grepl("^Area:", colnames(raw_data)) & !grepl("^Area: (Bl|QC)", colnames(raw_data)))) {
    stop(sprintf("The individual area columns are not present in the dataset in file '%s'", i))
  }
  
  # After verifying the correct file name, use everything between the first '_' and the file extension as the new file name. In other words, extract the part that comes after 'Compounds_'.
  if (!grepl("^compounds_", i, ignore.case = TRUE)) {
    stop(sprintf("Verify that the file '%s' has the appropiated name", i))
  }
  new_data_name <- sub(".*?_", "", i) %>% file_path_sans_ext()
  
  ## Change column names with the new name given by the correspondences dataset
  if (!"Additional_info" %in% colnames(correspondences)){
    for (n in 1:nrow(correspondences)) {
      current_name <- correspondences$Current_name[n]
      if (grepl("^area", current_name, ignore.case = TRUE)){
        new_name <- paste0("area_", correspondences$New_name[n])
      } else {
        new_name <- correspondences$New_name[n]
      }
      colnames(raw_data)[colnames(raw_data) %in% current_name] <- new_name
    }
  } else {
    for (n in 1:nrow(correspondences)) {
      current_name <- correspondences$Current_name[n]
      if (grepl("^area", current_name, ignore.case = TRUE)){
        new_name <- paste0("area_", correspondences$New_name[n], "_", correspondences$Additional_info[n])
      } else {
        new_name <- correspondences$New_name[n]
      }
      colnames(raw_data)[colnames(raw_data) %in% current_name] <- new_name
    }
  }
    
  # Substitute NA name metabolites with their formula or mz and RT values:
  for (n in 1:nrow(raw_data)) {
    if (is.na(raw_data$name[n])) {
      if (!is.na(raw_data$formula[n])) {
        raw_data$name[n] <- gsub("\\s", "", raw_data$formula[n])
      } else {
        raw_data$name[n] <- paste0("mz_", raw_data$mz[n], "_&_RT_", raw_data$RT[n])
      }
    }
  }
    
  # Sum areas of repeated metabolite and leave only a row for each one
  new_data <- raw_data %>% 
    group_by(name) %>%
    summarize(across(starts_with("Area"), ~sum(., na.rm = TRUE))) %>%
    ungroup()

  # Select the desired columns
  ## Select columns that begin with 'Area' but exclude 'Blank', 'QC', and 'Max'
  areas <- colnames(new_data)[grepl("^area_", colnames(new_data)) | grepl("^name", colnames(new_data))]
  new_data <- new_data[, areas]
  
  ## Add a new row with the group label
  group <- c(character(), "group")
  for (j in 1:nrow(correspondences)) {
    new_name <- correspondences$New_name[j]
    if (any(grepl(paste0("area_", new_name), colnames(new_data)))) {
      group <- c(group, correspondences$Group[j])
    }
  }
  new_data <- rbind(group, new_data)
  
  # Save data
  assign(new_data_name, new_data)
  
  # Save data names in a list for later merging
  env_files <- c(env_files, new_data_name)
  
  # Message of everything have work correctly
  message(sprintf("The initial file '%s' has been successfully edited and saved under the name '%s'.\n - Original dimensions: %d variables and %d observations\n - Updated dimensions: %d variables and %d observations\n - New observations name: %s", i, new_data_name, dim(raw_data)[1], dim(raw_data)[2], dim(new_data)[1], dim(new_data)[2], paste(colnames(new_data), collapse = ", ")))
  message(" ")
}
```


## Create dinamics tables:

```{r warning=FALSE}
# Save OTA and noOTA files in separate merged dataframes
OTA <- lapply(env_files[grepl("^(?!.*noOTA).*$", env_files, perl = TRUE)],
              function(file) get(file))
merged_OTA <- Reduce(function(df1, df2) merge(df1, df2, by = "name", all = TRUE), OTA)


noOTA <- lapply(env_files[grepl("noOTA", env_files)], function(file) get(file))
merged_noOTA <- Reduce(function(df1, df2) merge(df1, df2, by = "name", all = TRUE), noOTA)

env_files <- ls() %>% .[grepl("merged", ., perl = TRUE)]
for (i in env_files){
  merged_data <- get(i); name <- i
  # Move the group row information to the first position of the data frame
  merged_data <- rbind(merged_data[which(merged_data[, 1] == "group"), ],
                       merged_data[-which(merged_data[, 1] == "group"), ]) %>% `row.names<-`(NULL)
  # Eliminate those metabolites which are not in all the samples
  merged_data <- merged_data[complete.cases(merged_data), ]
  assign(name, merged_data)
}
  
# Merge the data frames, including the metabolites that are not common to both conditions
Global_merged_data <- merge(merged_noOTA, merged_OTA, by = "name", all = TRUE)
  
# Move the group row information to the first position of the data frame
Global_merged_data <- rbind(Global_merged_data[which(Global_merged_data[, 1] == "group"), ],
                            Global_merged_data[-which(Global_merged_data[, 1] == "group"), ]) %>% `row.names<-`(NULL)
  
# Set NA values as one-fifth of the minimum AUC
min_auc <- Global_merged_data[-1, -1] %>%
  mutate_all(as.numeric) %>%
  summarize(across(everything(), ~ min(.[. != 0], na.rm = TRUE), .names = "min_{.col}")) %>% 
  pull() %>% min(na.rm = TRUE)

Global_merged_data[is.na(Global_merged_data)] <- min_auc / 5

# Save merging
file_path <- file.path(output_folder, "Comparative_G.csv")
write.csv(Global_merged_data, file = file_path, row.names = FALSE)
        
# Print message with information about the merged data frames
message(paste("- File has bieen saved in", file_path, "."))

```
